---
title: "Gams and Non linear"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Library statements
library(dplyr)
library(readr)
library(broom)
library(ggplot2)
library(tidymodels)
tidymodels_prefer()

pitching1 = read.csv("stats.csv")



```

```{r}
pitching_data <- pitching1 %>%
  select(c(1,2,7,34,35,36,118:138)) %>%
  select(-c(1,2)) %>%
  select(!contains("formatted")) %>%
  drop_na()

gam_spec <-
  gen_additive_mod() %>%
  set_engine(engine = 'mgcv') %>%
  set_mode('regression')

fit_gam_model <- gam_spec %>%
  fit(p_k_percent ~ s(pitch_count_offspeed) + s(pitch_count_fastball) + s(pitch_count_breaking) + s(fastball_avg_speed) + s(fastball_avg_spin) + s(fastball_avg_break_x) + s(fastball_avg_break_z) + s(fastball_avg_break) + s(fastball_range_speed) + s(offspeed_avg_speed) + s(offspeed_avg_spin) + s(offspeed_avg_break_x) + s(offspeed_avg_break_z) + s(offspeed_avg_break) + s(offspeed_range_speed) + s(breaking_avg_speed) + s(breaking_avg_spin) + s(breaking_avg_break_x) + s(breaking_avg_break_z) + s(breaking_avg_break) + s(breaking_range_speed), data = pitching_data)
```

```{r}
# check to see if number of knots is good
par(mfrow=c(2,2))
fit_gam_model %>% pluck('fit') %>% mgcv::gam.check()
# I think this is good based on the in-class activity

# summary
fit_gam_model %>% pluck('fit') %>% summary()
```
Non-significant Coefficients: pcoff, pcfb, fbavgspd, fbavgbrkx, offavgspin, anything breaking - can be fixed

The coefficients with and exsimated degree of freedom of 1 have linear relationship. 

```{r}
fit_gam_model %>% pluck('fit') %>% plot( all.terms = TRUE, pages = 3)

```
Variables we want to fit to linear model. fastball_avg_speed, fastball_range-speed, offspeed_avg_speed, offspeed_avg_spin. 
Below we are creating a model that includes both splines and linear relationships with each variable 
```{r}
lm_spec <-
 linear_reg() %>%
 set_engine(engine = 'lm') %>%
 set_mode('regression')

set.seed(34) 
data_cv10 <- pitching_data %>%
 vfold_cv(v = 10)

pitching_rec <- recipe(p_k_percent ~ pitch_count_offspeed+ pitch_count_fastball + pitch_count_breaking + fastball_avg_speed + fastball_avg_spin + fastball_avg_break_x + fastball_avg_break_z + fastball_avg_break + fastball_range_speed + offspeed_avg_speed + offspeed_avg_spin + offspeed_avg_break_x + offspeed_avg_break_z + offspeed_avg_break + offspeed_range_speed + breaking_avg_speed + breaking_avg_spin + breaking_avg_break_x + breaking_avg_break_z + breaking_avg_break + breaking_range_speed, data=pitching_data)

# for all variables with edf over 3 we put them into our model as a spline 
spline_rec <- pitching_rec %>% 
  step_ns(fastball_avg_break_x, deg_free = 5)%>%
  step_ns(fastball_avg_break_z, deg_free=4)%>% 
  step_ns(offspeed_avg_break_x, deg_free=4)%>% 
  step_ns(offspeed_range_speed, deg_free=4)%>% 
  step_ns(breaking_range_speed,deg_free=4)

pitching_wf<- workflow()%>% 
  add_model(lm_spec)%>%
  add_recipe(pitching_rec)

spline_wf<- workflow()%>% 
  add_model(lm_spec)%>%
  add_recipe(spline_rec)

fit_resamples(
  pitching_wf, 
  resamples=data_cv10,
  metrics=metric_set(mae,rmse,rsq)
)%>% collect_metrics() 

fit_resamples(
  spline_wf, 
  resamples=data_cv10,
  metrics=metric_set(mae,rmse,rsq)
)%>% collect_metrics() 

  
```


