---
title: "hw4_trees"
author: "Jonah Klein-Collins"
date: "3/29/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(readr)
library(ggplot2)
library(dplyr)
library(tidymodels)
library(rvest)
tidymodels_prefer()
library(mosaic)
library(glmnet)
library(vip)
conflicted::conflict_prefer("vi", "vip")
load("pitching_clean.rdata")

```


```{r}
set.seed(123)

#model specification
rf_spec <- rand_forest()%>%
  set_engine(engine = 'ranger') %>%
  set_args(mtry = NULL,
           trees = 500, #Number of trees
           min_n = 2, #minimum number of points in a leaf?
           probability = FALSE,
           importance = 'impurity') %>%
  set_mode('classification')

#recipe
data_rec <- recipe(over25kperc ~ ., data = pitching_clean)

#workflows
data_wf_mtry5 <- workflow() %>%
  add_model(rf_spec %>% set_args(mtry = 5))%>%
  add_recipe(data_rec)
```

```{r}
# fit model
set.seed(123)
data_fit_mtry5 <- fit(data_wf_mtry5, data = pitching_clean)

rf_OOB_output <- function(fit_model, model_label, truth){
  tibble(
    .pred_class = fit_model %>% extract_fit_engine() %>% pluck('predictions'), 
    class = truth,
    label = model_label
  
  )
}

#check the output
rf_OOB_output(data_fit_mtry5,data_fit_mtry5$over25kperc, pitching_clean %>% pull(over25kperc))

# Evaluate OOB Metrics
data_rf_OOB_output <- bind_rows(
    rf_OOB_output(data_fit_mtry5,data_fit_mtry5$over25kperc, pitching_clean %>% pull(over25kperc)))

data_rf_OOB_output %>% 
  accuracy(truth = class, estimate = .pred_class) 
```
```{r}
data_fit_mtry10
rf_OOB_output(data_fit_mtry5, data_fit_mtry5$over25kperc, pitching_clean %>% pull(over25kperc)) %>%
  conf_mat(truth = class, estimate = .pred_class)

sensitivity = 74/(74+34)
specificity = 201/(201+14)
sensitivity
specificity
```
Our sensitivity is .685 and our specificity is .934. Therefore, our model is better at predicting when pitchers are below a 25% strikeout rate than above a 25% strikeout rate - it is better at predicting true negatives than true positives. We think this might be happening because pitchers that have a strikeout rate above 25% are very hard to come by - they are a somewhat elite group, even though the statistic is only slightly above the mlb average. Additionally, our data does not include relief pitchers, who tend to rely more on striking batters out than starting pitchers, who might rely on generating ground balls and soft contact.

```{r}
# evaluate predictor performance based on permutation
model_output <- data_wf_mtry5 %>% 
  update_model(rf_spec %>% set_args(importance = "permutation")) %>% #based on permutation
  fit(data = pitching_clean) %>% 
    extract_fit_engine() 

model_output %>% 
    vip(num_features = 30) + theme_classic()


model_output %>% vip::vi() %>% head()
model_output %>% vip::vi() %>% tail()
```
The most important variables in our model are the average speed,  spin, and break rate of a pitcher's fastball. This makes sense because this is a pitch that all pitchers typically use, and is important for pitchers who try to generate strikeouts by throwing the ball past the batter instead of trying to generate soft contact and ground balls.
